The stopping power of hydrogen in hafnium and \\ the importance of relativistic $4f$ electrons

### 2020-05-11 response.tex

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\setlength{\parindent}{0pt}
\usepackage{soul}
\usepackage{xcolor}
\usepackage[margin=0.8in]{geometry}
\usepackage{graphicx}

\setstcolor{red}
\def\reviewer#1{\vspace{0.35cm}\textsl{#1}}
\def\reply#1{\vspace{0.1cm}\textcolor{red}{#1}}

% Title Page
\title{Response to referee reports \\ AP11756/Montanari }
%\author{C. C. Montanari, A. M. P. Mendez, D. M. Mitnik, J. E. Miraglia, \\ P.A. Miranda, R. Correa, J. Wachter, M. Aguilera, \\ E. Alves, N. Catarino and R.C. da Silva}
%\date{\today}

@@

\begin{document}
\maketitle

% ----------------------------------------------------------------------
\section{Reply to report of the First Referee}
% ----------------------------------------------------------------------
\reviewer{1. Since only protons are considered as projectiles, the title 
needs to be modified accordingly.}

% {\color{red} The title has been modified following the referee's 
% recommendation:}{\small ''The stopping power of hydrgen in hafnium 
% and the importance of relativistic $4f$ electrons''}.
\reply{The title has been modified following the referee's
recommendation to ``The stopping power of hydrogen in hafnium and the
importance of relativistic $4f$ electrons''}.

% ----------------------------------------------------------------------
\reviewer{2. It appears that calculations assume bare protons. According 
to CasP, neutrals play a considerable role in charge equilibrium around 
and below the Bragg peak. Some explanation is needed here.}

%\reply{It is correct, our calculations assume bare protons. We consider that there is no bound electron or very small probability because free electrons of metals screens the hydrogen and this screening is enough at low energies as to have no bound electron. CasP includes a very good fitting of experimental charge states and a powerful scaling rule about them. However, it is based on measurements of charge states outside the metal. For multi-charged ions almost no differences are expected and CasP charge state works very well. For H and He ions, differences are important.  We add the following explanation about why considering just proton impact, and the reference to CasP and the paper of charge states by Schwietz and Grande.} \\ {\small ''Finally, in all our calculations we assumed the projectile to be proton and not neutral hydrogen. When an ion moves inside a metal, the FEG screens the nucleus, so the binding energies will be smaller than outside the metal, and this effect is more important a low impact velocities $v$. In the case of hydrogen the difference is drastic, i.e. for H inside Hf ($rs=2.07$), the $1s$-bound state is almost null at $v<2$. It is worth to mention that this assumption agrees with Ziegler SRIM code, but differs from CasP code that predicts neutral hydrogen at very low velocities.''}
\reply{The referee is correct; our calculations assume bare protons. We
considered that there is no bound electron (or insignificant probability
of there being any bound electron) because the free electrons in the
metal screen the hydrogen, and, at low energies, this screening is
sufficient as to have no bound electron. CasP includes an excellent
fitting of experimental charge states and a robust scaling rule about
them. However, it is based on measurements of charge states outside the
metal. For multi-charged ions, almost no differences are expected, and
the CasP charge state works very well. For H and He ions, differences
are significant.  We incorporated a new paragraph (lines 263-273) to
explain why we are considering just proton impact, the reference to
CasP, and the paper of charge states by Schwietz and Grande.}

% ----------------------------------------------------------------------
\reviewer{3. Thickness measurements by Rutherford backscattering rely 
‘heavily’ on the scattering cross section, while energy loss is a minor 
correction depending on foil thickness (p.1, 2nd paragraph).}

\reply{Pedro Miranda...................................................}

% ----------------------------------------------------------------------
\reviewer{4. In order to get a more precise thickness value of the Hf 
foil, the authors use the energy loss of 5.486 MeV alphas as a standard, 
with SRIM as a reference. SRIM is based on empirical data, and for alpha 
particles in Hf, the IAEA database lists one single data point for He 
in Hf. Considering the spread in published stopping cross sections for 
e.g. He in Au, listed in the IAEA database, some explanation is needed 
to justify an error estimate of only 5\% (1st paragraph in section 
IIB).}

%\reply{The referee is correct, we have used the stopping of 5.486 MeV alphas and SRIM as reference to get the thickness. Although SRIM fits the experimental data available, it is not just a fitting, it is a mix of model and fitting (semiempirical). At high impact energies where the stopping is well known srim is quite reliable. Figure \ref{HeHf} here can clarify these assertions. a) In this figure we plotted together H and He in Hf, dividing by $Z^2$. Above 1 MeV/amu both srim curves and our theoretical results agree within $2\%$. The vertical pink line indicates the impact energy we use as reference. It is in the region where the srim predictions are reliable. b) srim is semiempirical, the comparison of H and He in the same target shows clearly how srim for H ions curves down and up to fit Sirotonin data. It is also clear how srim for He ion overvalue in order to fit the single measurement by Chu, Ziegler and others. Why do we say "overvalue"? Because the stopping of He ions divided by 4 can be equal or below the stopping of H ions. }
\reply{The referee is correct; we have used the stopping of 5.486 MeV
alphas and the SRIM software package as a reference to get the 
thickness. Although SRIM fits the experimental data available, it is not
just a fitting; the software implements a mix of modeling and fitting
(semiempirical). At high impact energies, where the stopping is well
known, SRIM is quite reliable. Figure \ref{HeHf} included in this
reply can clarify these assertions. a) In this figure, we plotted H and 
He in Hf, both divided by $Z^2$. Above 1 MeV/amu, SRIM and our
theoretical results agree within $2\%$. The vertical pink line indicates
the impact energy used as reference. In the region, the SRIM predictions
are reliable. b) By comparing the SRIM results for hydrogen and helium
in the same target, one can see that SRIM for H ions rapidly changes its
concavity to fit Sirotonin data. Furthermore, SRIM for He ion 
overestimates the scaled-with-$Z^2$ cross-section at 500 keV/amu in
order to fit the single measurement by Chu, Ziegler, and collaborators. 
We say the stopping is overestimated because the stopping of He ions 
divided $Z^2$ can only be equal or below the stopping of H ions.}

%-----------------------------------------------------------------------
\begin{figure}[!t]
\centering
\includegraphics[width=12.cm]{Fig_HeHf.eps}
\caption{Stopping cross section of H and He in hafnium at high impact energies.}
\label{HeHf}
\end{figure}

%-----------------------------------------------------------------------
\reviewer{5. In connection with the discussion of relativity and 
outer-shell electrons (p.3) it might be instructive to mention that 
this is the reason for the existence of relativistic quantum chemistry
programs. Since the authors find it surprising that the effect increases
in magnitude from inner to outer shells, the reader needs an
explanation. I should think the effect of deviations in the screening
add up shell by shell, but I may be wrong. Moreover, the sign of the
discrepancy in figure 2a deserves attention, which seems to change from
positive up to 4p- to negative from 5p+.}

%\reply{We included two new paragraphs (lines 211-230) to make reference to the referee's comments on the binding energy calculations.}
\reply{The referee is correct. We included two paragraphs in Section III
(lines 211-230) to refer to her/his comments.}

% ----------------------------------------------------------------------
\reviewer{6. The importance of 4f-5p screening is documented clearly 
(p.3), but the reader will ask why this does not affect other shells. 
I assume that the effect is important when two or more subshells have
similar energies and, hence, overlap in real and velocity space. If so,
I suggest to mention this. If not, another explanation is needed.}

%\reply{The mixing criteria is explained on page 3: we consider the quantum uncertainty in energy so bound electrons with similar energies and mean radio $<r>$ response as a single density of electrons including screening among them. We recognize that a more detailed study is needed, we are developing such work as a systematic for different targets involving 4f electrons. What we have done here for Hf is to check the importance of the contribution to the total stopping. Only the 5p-4f mixing have influence in the total stopping because they are the second most important contribution after the FEG. Some other shells may be mixed, and we calculate them, but the total stopping changes in very small $\%$.  We added lines 262-264 to page 4.} {\small '' As already mentioned, at higher energies inter-shell screening is possible for other subshells (i.e. 4p-4d for impact energies above 0.9 MeV), but its weight in the total stopping is minor for deeper shells.'' }
\reply{The mixing criterion is explained on page 3; we considered
the quantum uncertainty in energy so that bound electrons with similar
energy values and mean radio $\langle r\rangle$ response as a single
density of electrons, including screening among them. We acknowledge
that a more detailed study is required. We are currently working on
developing a systematic approach for different targets involving 4f
electrons. So far, we have corroborated the importance of the
contribution of the 4f shell to the total stopping in Hf. We have also
determined that the 5p-4f mixing influences the total cross section
mainly because they are the second most significant contribution after
the FEG. Other shells may be mixed, and we have computed them, but the
total stopping varies very little. Reference to this analysis is given
in lines 258-262.}

% ----------------------------------------------------------------------
\reviewer{7. In view of the large contribution of the outermost shell, 
I suggest to include the core contribution in figure 3.}

%\reply{We thank the referee for this suggestion, actually we had this figure, but we doubted on including all the curves. Now Fig. 3 includes separate FEG and bound electron contribution, and the total curves too.}
\reply{We thank the referee for her/his suggestion. At first, we thought 
of showing all contributions separately but opted for a simpler version.
Upon recommendation, we have included in Fig. 3 of the manuscript the
FEG and bound electron contribution curves.}

% ----------------------------------------------------------------------
\reviewer{8. In addition to comparisons with experiments and empirical 
models, it would be instructive to see comparisons with theoretical 
models like CasP and DPASS in figure 4.}

%\reply{We thank this suggestion. We had already compared with Casp5.2 and DPASS for ourselves before submitting. In this version we added CasP5.2 and DPASS to figure 4, we added a mention in the introduction(last paragraph, lines 68-73), and the following comment in page 6, lines 315-322:} {\small ''The stopping maximum is a very sensitive region for any full theoretical description, with no parameters at all, and this is quite visible in a linear-scale plot like Fig. [4]. However, the impact energy for the maximum seem to agree between our curve and DPASS, though $10 \%$ below. Instead, CasP predicts a stopping maximum $10 \%$ above and a completely different shape at lower energies.''}
\reply{We compared our results with Casp5.2 and DPASS before submission,
but we were discouraged to include them by the number of curves in the
figure. However, we have now managed to include CasP5.2 and DPASS to
figure 4. We comment about these changes in the Introduction 
(lines 71-72), and in Section IV (lines 308-311).}

\newpage
% ----------------------------------------------------------------------
\section{Reply to report of the Second Referee}
% ----------------------------------------------------------------------
\reviewer{1. Can the authors discuss how possible impurities in the 
sample and -especially- on both sides of the target would affect the 
final stopping data? The authors must discuss it in the manuscript.}

\reply{Pedro Miranda...................................................}

% ----------------------------------------------------------------------
\reviewer{2. In the same context, the authors have to clarify whether
potential surface roughness and/or target non-uniformity might affect 
the final results obtained in transmission method.}

\reply{Pedro Miranda...................................................}

% ----------------------------------------------------------------------
\reviewer{3. What is accuracy of the primary beam energy of the 
accelerator? Moreover: when the target is placed and removed from the 
front of the detector, does the detector itself need to unbiased and the 
chamber opened? If so, might there be any offset ($\sim$keV) in the 
detection system that would affect the energy distribution peaks, hence 
the Gaussian position from the fits (Fig. 1)?}

\reply{Pedro Miranda...................................................}

% ----------------------------------------------------------------------
\reviewer{4. The experimental data above $\sim$1.5 MeV is systematically 
below the data from Ref. [7], hence SRIM semi-empirical approach, and 
also below the authors’ theory output. At these energies (and higher), 
the agreement between experiment and theory is expected to be better 
than a few percent (that is why you used stopping power for 5.486 MeV 
alpha on Hf to obtain the target thickness). Can the authors explain why
there is such systematic difference for the present measurements using
protons?}

% \reply{There is a small systematic difference between our experiments and theory for impact energies above 1.5 MeV, but it is within the experimental error. Figure 2 of this reply amplifies the results shown in the manuscript by focusing on the high energy region. Note that we included our theoretical results, SRIM, Casp5.2, and DPASS in the figure. The difference is small, being less than 3\% for the two values above 2 MeV. We included comments regarding this subject in the manuscript. <=== Pedro, PLEASE CHECK THIS}
\reply{There is a small systematic difference between our experiments and theory for impact energies above 1.5 MeV, but it is within the experimental error. Figure 2 of this reply amplifies the results shown in the manuscript by focusing on the high energy region. Note that we included our theoretical results, SRIM, Casp5.2, and DPASS in the figure. The difference is small, being less than 3\% for the two values above 2 MeV. We included comments regarding this subject in the manuscript.}

%-----------------------------------------------------------------------
\begin{figure}[!t]
\centering
\includegraphics[width=12.cm]{Fig04.eps}
\caption{Stopping at high impact energies. Theoretical curves and
symbols for data are explained inside the figure.}
\label{highE}
\end{figure}

%-----------------------------------------------------------------------
\reviewer{5. The authors claim an overall uncertainty of 5\%, but it is 
not clear how they obtained this value, or whether only statistical
contributions were taken into account. Nevertheless, uncertainties of
measurements have to be treated following the Guide to the Expression
of Uncertainty in Measurement, International Organization for
Standardization, Geneva Switzerland, 1995. Especially for stopping
power measurements in transmission approach, there are other sources
of uncertainties that might potentially affect the final results
systematically (see comments 1, 2 and 3), and they must be discussed
in more details.}

\reply{Pedro Miranda...................................................}

% ----------------------------------------------------------------------
\reviewer{6. Can the present theory in the manuscript be compared to 
other models (e.g., CasP and DPASS), especially towards low energies? 
Aiming thus a constructive discussion on the overall-agreement among
different models to predict SCS of “challenge” materials, as transition 
metals (f-subshell)?}

% \reply{We thank the referee for this suggestion that improves this new version. We have added CasP5.2 and DPASS curves in Figure 4, we have also added a mention in the introduction (last paragraph, lines 68-73), and the following comment in page 6, lines 315-322:} \\ {\small ''The stopping maximum is a very sensitive region for any full theoretical description, with no parameters at all, and this is quite visible in a linear-scale plot like Fig. [4]. However, the impact energy for the maximum seem to agree between our curve and DPASS, though $10 \%$ below. Instead, CasP predicts a stopping maximum $10 \%$ above and a completely different shape at lower energies.''}
\reply{We thank the referee for her/his suggestion. We have included the 
CasP5.2 and DPASS curves in Fig. 4 of the manuscript. We made the 
corresponding modification in the Introduction (lines 71-72), and 
Section IV (lines 308-311).}

% ----------------------------------------------------------------------
\reviewer{7. In the last sentence of Sec. IV, the authors point out for 
a need of more experimental data below 100 keV H+ but, in my opinion, 
more data are needed even at higher energies, e.g., nearby (and slightly
above) the Bragg peak... Please enhance discussions about it.}

% \reply{We agree with this comment by the referee. In a ion-target system with only two set of measurements more data is needed. In the energy region 80-500 keV only the data by Sirotinin group is available and cannot be compared with others. SRIM is not a test for this data, SRIM includes this data in the adjust (more detail about this can be found in the answer 4 to the first referee report). We change the comments in page 6 lines 328-336 and lines 364-366 as follows:} \\ {\small ''Future experiments  would be important to complete this study on the stopping of Hf for protons. In the energy region $80-500$ keV only the measurements by Sirotinin group in 1984 are available. SRIM clearly includes these values in the adjust of the code, so it does not represent a test for this data. Undoubtedly, new measurements are needed in three specific energy regions: below 500 keV, around the stopping maximum (i.e. $50-200$ keV) and at low impact energies''} \\ {\small ''Future experiments for impact energies below 500 keV, around the stopping maximum and in the low energy region would be important to complete this study.''}
\reply{We agree with the referee's comment. In an ion-target system with only two sets of measurements, more data is needed. In the energy region 80-500 keV, only the data by Sirotinin et al. is available and cannot be compared with others. SRIM is not a test for this data since it adjusts these measurements itself (a more detailed discussion about the matter can be found in reply 4 of the first referee report). We changed our manuscript accordingly in Section IV (lines 331-339) and Section V (lines 374-377).}

\newpage
% ----------------------------------------------------------------------
\section{Reply to report of the Third Referee}
% ----------------------------------------------------------------------
\reviewer{1. Regarding the theoretical model that describes the
contribution to the stopping power of the bound electrons, the authors
emphasize the importance of including relativistic effects in the
description of the binding energies and densities of the Hf 1s-4f
states. However, it is not clear if the calculations performed by the
authors are done for gas phase Hf atoms or for the solid Hf as in the
experiments of Ref. [38]. In case of being for the former, the slightly
worse agreement found for the 4f states as compared to the experimental
values could be related to this point.}

%\reply{The referee is correct, our calculations are performed for gas phase Hf atoms. We have explicitly clarify the difference between the experimental and theoretical binding energies in page 3, lines 215-219.}
\reply{The referee is correct; our calculations are performed for gas-
phase Hf atoms. We have now explicitly clarified the difference between 
the experimental and theoretical binding energies in the manuscript 
(lines 215-219).}

% ----------------------------------------------------------------------
\reviewer{2. The results from the three different theoretical
calculations shown in Fig. 3 are not completely clear. The authors
emphasize the importance of considering both screening effects and
relativistic effects for the 4f and 5p electrons, but it is not clear if
later in the text the authors use indistinctly the terminology
``relativistic'' and ``screening'' to refer to the same issue. For
instance, are the ML results shown in Fig. 3 calculated with the
relativistic binding energies and densities, but with and without
including screening corrections, or do the two ML curves differ in
whether the relativistic corrections are or not included? It is neither
clear what description of the bound electrons is used in the curve
labeled as ``SPCC(FEG)+SLPA(bound)''.}

### 2020-05-11 sp_hafnium.tex
checked english with grammarly

%Title of paper
\title{Stopping power of hafnium by proton impact and \\
the importance of relativistic $4f$ electrons}
%Authors and affiliations

%---------------------------Argentina
\author{C. C. Montanari}
\email{mclaudia@iafe.uba.ar}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%------------------------------
\author{A. M. P. Mendez}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%-------------------------------
\author{D. M. Mitnik}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%----------------------------------------------
\author{J. E. Miraglia}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%---------------------------Chile

\author{P.A. Miranda}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%----------------------------
\author{R. Correa}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%----------------------------
\author{J. Wachter}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%----------------------------
\author{M. Aguilera}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%-------------------------Lisboa
\author{E. Alves}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}
\affiliation{Instituto de Plasma e Fus\~{a}o Nuclear, Instituto Superior 
T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, Portugal}
%---------------------------
\author{N. Catarino}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}
%---------------------------
\author{R.C. da Silva}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}

@@

%=======================================================================
%=======================================================================
\begin{abstract}
 The stopping power of protons through Hf foil has been studied both 
experimentally and theoretically. The measurements were performed at 
the Laboratory of Accelerators and X-Ray Diffraction in Lisbon, by 
using the transmission method on self-supporting stopping material. 
 The overall uncertainty of around $5\%$ was established over the 
protons energy range  (0.6-2.5) MeV. The theoretical developments 
involved fully relativistic atomic structure calculations for Hf, which 
required the solution of the Dirac equation. The shell-wise local plasma 
approximation (SLPA) was used to describe the energy transferred to the 
bound $1s$-$4f$ electrons, and the outer four electrons were considered 
as a free electron gas (FEG). We found the relativistic description of 
the $4f$-shell and the screening between $4f$ and $5p$ electrons are 
decisive around the stopping maximum. Present theoretical and 
experimental results are in very good agreement in the energy region 
of the new measurements. However, the present theoretical stopping 
cross sections show substantial differences with the most used 
semi-empirical models (SRIM2013 and ICRU-49) at intermediate to low 
energies. Our calculations suggest the stopping maximum higher and 
shifted to lower energies than these previous predictions. Future 
measurements below 100 keV would be decisive for a better knowledge 
around the maximum and below.
\end{abstract}

@@

For impact energies above a few keV/amu, mono-energetic charged 
particles penetrating a foil of any material lose their energy through 
a series of consecutive inelastic collisions, mainly with target 
electrons~\cite{Chu01,Sigmund}. The information given by the energy 
loss process is essential not only to have a better knowledge of the 
physics behind the fundamental interactions, but also because it plays 
an important role in many applied fields such as materials science, 
nuclear physics, ionic implantation and radiotherapy~\cite{Sigmund,Schardt}. 
Experimental data on ion mean energy loss per unit path $S(E)$ is of 
crucial relevance to check the reliability of semi-empirical models and 
to determine some key parameters~\cite{Diwan,Damache04,Damache02}. The 
experimental data available is often rather scarce, which is troublesome 
when the material under study corresponds to an element of low 
occurrence on the Earth's upper crust, such as hafnium.

So far, only one experimental work has been published regarding the 
stopping power cross section of pure hafnium for protons~\cite{Sirotinin}, 
while more attention has been recently given to studies involving 
hafnium oxide due to its practical use~\cite{Abril,Behar,Primetzhofer,Roth}. 
It is well known that significant attention has been paid in recent 
years to transition metal-oxides such as HfO$_2$ because of their 
potential as alternative gate dielectrics to replace SiO$_2$ for the 
future generation of nano-electronics with less than 45 nm gate 
length~\cite{Choi,Robertson}. Some important physical properties of the 
above mentioned metal-oxide films depend on its thickness, which is 
often measured by using Rutherford Backscattering Spectrometry~\cite{Alfassi01,Tesmer01}, 
a method that relies heavily on the determination of the stopping power 
of ion beams in the material of interest.

In this study, we report experimental stopping power cross sections 
over the incident energy range (0.6-2.5) MeV for protons crossing 
self-supported Hf thin-film by using the transmission method. We aim 
not only to upgrade stopping power data compilations~\cite{HPaul03,mondim17} 
but also to provide useful information about the processes governing 
the slowing down of protons in multi-electronic targets. In the rare 
earth metals, the $4f$ electrons play a relevant role in the stopping 
power, being the first shell of bound electrons below the conduction 
band. As already noted \cite{Roth17}, the free electron gas (FEG) shows 
unexpected behavior in these elements, which casts doubts on its proper 
description. In the case of Hf, we found the contribution of the 
$4f$-shell decisive even at impact energies around the stopping maximum, 
as shown later.

The theoretical approach implemented in this work uses the shell-wise 
local plasma approximation (SLPA)~\cite{mon13} to describe the energy 
transferred to the bound $1s$-$4f$ electrons, and two different models 
for the FEG: the screened potential with cusp condition model (SPCC)~\cite{mon17}, 
which is non-linear binary formalism; and the Mermin-Lindhard dielectric 
formalism (ML)~\cite{Mermin}, for energies around the stopping maximum 
and above. Our model requires the relativistic wave functions and 
binding energies of Hf, and considers 4 electrons per atom in the 
FEG~\cite{mendez2019}. Hf has the extra interest of the filled 
$4f$-subshell (with 14 electrons) as the main contribution below the 
FEG, causing the stopping cross sections to be very sensitive to a good 
description of this shell. The screening among the $4f$ and $5p$ 
electrons has been considered and found to play a major role within the 
SLPA calculations.

The experimental details and data are given in Section~\ref{experiment}, 
while the theoretical method is explained in Section~\ref{theory}. 
Present theoretical and experimental values are finally compared to the 
available experimental values~\cite{Sirotinin}, the ICRU-49 
calculations~\cite{ICRU49} and the SRIM-2013 package~\cite{Ziegler01}. 
Conclusions and discussions are given in Section~\ref{conclusion}. All 
the present data can be found at the Zenodo platform \cite{zenodo}.

%-----------------------------------------------------------------------
\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Fig01.eps}
\caption{RBS spectrum for $E_{\mathrm{avg}}=921.1$ keV protons on 
hafnium sample which is subsequently used to determine the energy loss 
in the foil.}
\label{F01}
\end{figure}

@@

The present data and theoretical results are displayed in 
Table~\ref{table01}, and Fig.~\ref{F03}. As can be seen in 
Table~\ref{table01}, an overall relative uncertainty of around 5\% was 
achieved for the experimental stopping power values, which are mainly 
due to the uncertainty in the hafnium foil thickness.

>>>>

%Title of paper
\title{The stopping power of hydrogen in hafnium and \\
the importance of relativistic $4f$ electrons}
%Authors and affiliations

%---------------------------Argentina
\author{C. C. Montanari}
\email{mclaudia@iafe.uba.ar}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}

%---------------------------Chile
\author{P. A. Miranda}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%-------------------------Lisboa
\author{E. Alves}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}
\affiliation{Instituto de Plasma e Fus\~{a}o Nuclear, Instituto Superior 
T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, Portugal}

%------------------------------Argentina
\author{A. M. P. Mendez}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%-------------------------------
\author{D. M. Mitnik}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}
%----------------------------------------------
\author{J. E. Miraglia}
\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio, 
Consejo Nacional de Investigaciones Cient\'{\i}ficas y T\'{e}cnicas - 
Universidad de Buenos Aires, Pabell\'on IAFE, 1428 Buenos Aires, 
Argentina}

%---------------------------- Chile
\author{R. Correa}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%----------------------------
\author{J. Wachter}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}
%----------------------------
\author{M. Aguilera}
\affiliation{Departamento de F\'isica, Facultad de Ciencias Naturales, 
Matem\'atica y del Medio Ambiente. Universidad Tecnol\'ogica Metropolitana. 
7800002, Chile}

%-------------------------Lisboa
\author{N. Catarino}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}
%---------------------------
\author{R. C. da Silva}
\affiliation{Centro de Ci\^{e}ncias e Tecnologias Nucleares, Instituto 
Superior T\'ecnico, Universidade de Lisboa, 2696-953 Sacav\'{e}m, 
Portugal}

@@

%=======================================================================
%=======================================================================
\begin{abstract}
 The stopping power of protons through Hf foil has been studied both 
experimentally and theoretically. The measurements were performed at 
the Laboratory of Accelerators and X-Ray Diffraction in Lisbon by 
using the transmission method on self-supporting stopping material. 
 The overall uncertainty of around $5\%$ was established over the 
protons energy range  (0.6-2.5) MeV. The theoretical developments 
involved fully relativistic atomic structure calculations for Hf, which 
required the solution of the Dirac equation. The shell-wise local plasma 
approximation (SLPA) was used to describe the energy transferred to the 
bound $1s$-$4f$ electrons, and the outer four electrons were considered 
as a free electron gas (FEG). We found the relativistic description of 
the $4f$-shell and the screening between $4f$ and $5p$ electrons to be 
decisive around the stopping maximum. Present theoretical and 
experimental results are in excellent agreement in the energy region 
of the new measurements. However, our theoretical stopping 
cross sections show substantial differences with the most used 
semi-empirical models (SRIM2013 and ICRU-49) at intermediate to low 
energies. Our calculations suggest the stopping maximum to be higher and 
shifted to lower energies than these previous predictions. Future 
measurements below 100 keV would be necessary for a better understanding 
of the stopping power around the maximum and below.
\end{abstract}

@@

For impact energies above a few keV/amu, mono-energetic charged 
particles penetrating a foil of any material lose their energy through 
a series of consecutive inelastic collisions, mainly with target 
electrons~\cite{Chu01,Sigmund}. The information given by the energy 
loss process is essential not only to have a better comprehension of the 
physics behind the fundamental interactions but also because it plays 
a vital role in many applied fields such as materials science, 
nuclear physics, ionic implantation, and radiotherapy~\cite{Sigmund,Schardt}. 
Experimental data on ion mean energy loss per unit path $S(E)$ is of 
crucial relevance to check the reliability of semi-empirical models and 
to determine some key parameters~\cite{Diwan,Damache04,Damache02}. The 
experimental data available is often rather scarce, which is troublesome 
when the material under study corresponds to an element of low 
occurrence on the Earth's upper crust, such as hafnium.

So far, only one experimental work has been published regarding the 
stopping power cross section of pure hafnium for protons~\cite{Sirotinin}, 
while more attention has been recently given to studies involving 
hafnium oxide due to its practical use~\cite{Abril,Behar,Primetzhofer,Roth}. 
It is well known that significant attention has been paid in recent 
years to transition metal-oxides such as HfO$_2$ because of their 
potential as alternative gate dielectrics to replace SiO$_2$ for the 
future generation of nano-electronics with less than 45 nm gate 
length~\cite{Choi,Robertson}. Some important physical properties of the 
above mentioned metal-oxide films depend on their thickness, which is 
often measured by using Rutherford Backscattering 
Spectrometry~\cite{Alfassi01,Tesmer01}. This method relies heavily on 
the determination of the stopping power of ion beams in the material of 
interest.

In this study, we report experimental stopping power cross sections over 
the incident energy range (0.6-2.5) MeV for protons crossing 
self-supported Hf thin-film by using the transmission method. We aim not 
only to upgrade stopping power data compilations~\cite{HPaul03,mondim17} 
but also to provide useful information about the processes governing the 
slowing down of protons in multi-electronic targets. In the rare earth 
metals, the $4f$ electrons play an essential role in the stopping power 
since they belong to the first shell of bound electrons below the 
conduction band. As already noted \cite{Roth17}, the free electron gas 
(FEG) shows unexpected behavior in these elements, which casts doubts 
on its proper description. In the case of Hf, we found the contribution 
of the $4f$-shell to be decisive even at impact energies around the 
stopping maximum, as will be shown later.

The theoretical approach implemented in this work uses the shell-wise 
local plasma approximation (SLPA)~\cite{mon13} to describe the energy 
transferred to the bound $1s$-$4f$ electrons and two different models 
for the FEG; in low energy region, the screened potential with cusp 
condition model (SPCC)~\cite{mon17}, which is non-linear binary 
formalism, and the Mermin-Lindhard dielectric formalism 
(ML)~\cite{Mermin} for energies around the stopping maximum and above. 
Our model requires the relativistic wave functions and binding energies 
of Hf and considers four electrons per atom in the FEG~\cite{mendez2019}. 
Hafnium is particularly interesting since the filled $4f$-subshell (with 
14 electrons) is the main contributor below the FEG, causing the 
stopping cross sections to be very sensitive to a good description of 
this shell. The screening among the $4f$ and $5p$ electrons has been 
considered and found to play a significant role within the SLPA 
calculations.

The experimental details and data are given in Section~\ref{experiment}, 
while the theoretical method is explained in Section~\ref{theory}. 
Present theoretical and experimental values are finally compared to the 
only experimental values measured by Sirotinin {\it et al.}~\cite{Sirotinin}  
with the backscattering method, the theoretical results by Grande and 
Schiwietz~\cite{Grande,casp52}, and by Sigmund and Schinner~\cite{DPASS20}, 
and also with the semi-empirical values from the SRIM-2013 
package~\cite{Ziegler01} and the ICRU-49 tabulation~\cite{ICRU49}. 
Conclusions and discussions are given in Section~\ref{conclusion}. All 
the present data can be found at the Zenodo platform \cite{zenodo}.

%-----------------------------------------------------------------------
\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Fig01.eps}
\caption{RBS spectrum for $E_{\mathrm{avg}}=921.1$ keV protons on 
the hafnium sample, which is subsequently used to determine the energy 
loss in the foil.}
\label{F01}
\end{figure}

@@

The present data are displayed in Table~\ref{table01}. As can be 
observed, an overall relative uncertainty 
of around 5\% was achieved for the experimental stopping power values, 
which are mainly due to the uncertainty in the hafnium foil thickness.

%-----------------------------------------------------------------------
\begin{table*}[!t]
\centering
\caption{Stopping power values S$_{\mathrm{exp}}$ of hafnium for protons 
measured in this work. $\Delta$E/E values are also shown.}
\label{table01}
